---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 5 - Due date 02/13/24"
author: "Aditi Jackson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A05_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

R packages needed for this assignment: "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r packages}
#Load/install required package here
#install.packages("readxl")
library(readxl)
library(forecast)
library(tseries)
library(ggplot2)
library(Kendall)
library(lubridate)
library(scales)
library(cowplot)
library(tidyverse)  #load this package so you clean the data frame using pipes

```

## Decomposing Time Series

Consider the same data you used for A04 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2023 Monthly Energy Review.

```{r importing data}
# loading data using read.csv; note that I tried readxl, but it kept producing errors
energy_data <- 
  read.csv(
  "/home/guest/ENV797_APJ_S24_NEW/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.csv",
  header = TRUE, dec = ".", sep=",",stringsAsFactors = TRUE)

# renaming the "Month" column to "Date"
colnames(energy_data)[colnames(energy_data) == "Month"] <- "Date_MY"

# converting the "Date" column to a date object using lubridate
energy_data$Date_MY <- ym(energy_data$Date_MY)

# Verifying data
head(energy_data)

# creating variables to represent # of rows & # of variables in data
nobs=nrow(energy_data)
nvar=ncol(energy_data)

```

### Q1

For this assignment you will work only with the following columns: Solar Energy Consumption and Wind Energy Consumption. Create a data frame structure with these two time series only and the Date column. Drop the rows with *Not Available* and convert the columns to numeric. You can use filtering to eliminate the initial rows or convert to numeric and then use the drop_na() function. If you are familiar with pipes for data wrangling, try using it!

```{r subsetting data}
# subsetting data into data frame with three variables: Date, Solar Energy Consumption
## and Wind Energy Consumption
# using mutate to convert solar and wind columns to numeric
# using drop_na() to take care of any N/A values
energy_data_sub <- energy_data %>%
  select(
    Date_MY,
    Solar.Energy.Consumption,
    Wind.Energy.Consumption) %>%
  mutate(
    Solar.Energy.Consumption = as.numeric(Solar.Energy.Consumption),
    Wind.Energy.Consumption = as.numeric(Wind.Energy.Consumption))%>%
  drop_na()
  
```

### Q2

Plot the Solar and Wind energy consumption over time using ggplot. Plot each series on a separate graph. No need to add legend. Add informative names to the y axis using `ylab()`. Explore the function `scale_x_date()` on ggplot and see if you can change the x axis to improve your plot. Hint: use `scale_x_date(date_breaks = "5 years", date_labels = ("%Y")")`

```{r initial time series plots}
# plotting solar
ggplot(energy_data_sub,aes(x=Date_MY,y=Solar.Energy.Consumption))+
  geom_line()+
  ylab("Solar Energy Consumption (Trillion BTU)")+
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y"))+
  xlab("Years")

# plotting wind
ggplot(energy_data_sub,aes(x=Date_MY,y=Wind.Energy.Consumption))+
  geom_line()+
  ylab("Wind Energy Consumption (Trillion BTU)")+
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y"))+
  xlab("Years")

```

### Q3

Now plot both series in the same graph, also using ggplot(). Use function `scale_color_manual()` to manually add a legend to ggplot. Make the solar energy consumption red and wind energy consumption blue. Add informative name to the y axis using `ylab("Energy Consumption)`. And use function `scale_x_date()` to set x axis breaks every 5 years.

```{r initial time series plots cont'd}

# plotting solar & wind together
ggplot(energy_data_sub,aes(x=Date_MY))+
  geom_line(aes(y=Solar.Energy.Consumption,color="Solar"))+
  geom_line(aes(y=Wind.Energy.Consumption,color="Wind"))+
  scale_color_manual(values = c("Solar" = "red", "Wind" = "blue"))+
   theme(legend.title = element_blank(), legend.position = "bottom")+
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y"))+
  labs(x="Years",y="Energy Consumption (Trillion BTU)")
 
```


## Decomposing the time series

The stats package has a function called decompose(). This function only takes time series object. As the name says the decompose function will decompose your time series into three components: trend, seasonal and random. This is similar to what we did in the previous script, but in a more automated way. The random component is the time series without seasonal and trend component. 

Additional info on `decompose()`.

1) You have two options: alternative and multiplicative. Multiplicative models exhibit a change in frequency over time.
2) The trend is not a straight line because it uses a moving average method to detect trend. 
3) The seasonal component of the time series is found by subtracting the trend component from the original data then grouping the results by month and averaging them. 
4) The random component, also referred to as the noise component, is composed of all the leftover signal which is not explained by the combination of the trend and seasonal component. 

### Q4

Transform wind and solar series into a time series object and apply the decompose function on them using the additive option, i.e., `decompose(ts_data, type = "additive")`. What can you say about the trend component? What about the random component? Does the random component look random? Or does it appear to still have some seasonality on it?

```{r decomposing time series, type = additive}
# creating time series objects fo solar and wind
ts_solar = ts(energy_data_sub[,2],start=c(1973,1),frequency=12)
ts_wind = ts(energy_data_sub[,3],start=c(1973,1),frequency=12)

# decomposing solar and wind using additive type
decomp_add_solar <- decompose(ts_solar, type = "additive")
decomp_add_wind <- decompose(ts_wind, type = "additive")

# plotting decomposed solar and wind objects
autoplot(decomp_add_solar)+
  ggtitle("Decomposition of additive time series - SOLAR")
autoplot(decomp_add_wind)+
    ggtitle("Decomposition of additive time series - WIND")

```

> The trend component of solar shows a generally increasing trend. The trend component of wind shows a generally increasing trend. The random components of both series appear to still have some seasonality present. This is because both random components have a pretty constant sinusoidal pattern. [REVISE WIND]


### Q5

Use the decompose function again but now change the type of the seasonal component from additive to multiplicative. What happened to the random component this time?

```{r decomposing time series, type = multiplicative}
# decomposing solar and wind using multiplicative type
decomp_mult_solar <- decompose(ts_solar, type = "multiplicative")
decomp_mult_wind <- decompose(ts_wind, type = "multiplicative")

# plotting decomposed solar and wind objects
autoplot(decomp_mult_solar)+
    ggtitle("Decomposition of multiplicative time series - SOLAR")
autoplot(decomp_mult_wind)+
 ggtitle("Decomposition of multiplicative time series - WIND")

# plotting everything together to see changes
plot_grid(autoplot(decomp_add_solar),
          autoplot(decomp_add_wind),
          autoplot(decomp_mult_solar),
          autoplot(decomp_mult_wind))

```

> It appears that the random component of solar decreased in magnitude, and the general wave pattern has been mostly eliminated. For wind, [ADD] 

### Q6

When fitting a model to this data, do you think you need all the historical data? Think about the data from 90s and early 20s. Are there any information from those years we might need to forecast the next six months of Solar and/or Wind consumption. Explain your response.

> Answer: [ADD]

### Q7

Create a new time series object where historical data starts on January 2012. Hint: use `filter()` function so that you don't need to point to row numbers, .i.e, `filter(xxxx, year(Date) >= 2012 )`. Apply the decompose function `type=additive` to this new time series. Comment the results. Does the random component look random? Think about our discussion in class about seasonal components that depends on the level of the series. 

```{r subsetting data again}

# filtering data for Jan 2022 onwards
energy_data_sub2 <- energy_data_sub %>%
  filter(year(Date_MY)>=2012)

# creating new ts objects for solar and wind
ts_solar2 <- ts(energy_data_sub2[,2],start=c(2012,1),frequency=12)
ts_wind2 <- ts(energy_data_sub2[,3],start=c(2012,1),frequency=12)

# creating new decompositions
decomp_add_solar2 <- decompose(ts_solar2, type = "additive")
decomp_add_wind2 <- decompose(ts_wind2, type = "additive")

# plotting decomposed time series
autoplot(decomp_add_solar2)+
ggtitle("Decomposition of additive time series - SOLAR")

autoplot(decomp_add_wind2)+
   ggtitle("Decomposition of additive time series - WIND")

```

> Answer: [ADD]

## Identify and Remove outliers

### Q8

Apply the `tsclean()` to both series from Q4. Did the function removed any outliers from the series? Hint: Use `autoplot()` to check if there is difference between cleaned series and original series.

```{r removing outliers}

# removing outliers with tsclean()
solar_clean <- tsclean(ts_solar)
wind_clean <- tsclean(ts_wind)

# plotting "cleaned" series
autoplot(solar_clean)
autoplot(wind_clean)

```

> ADD EXPLANATION

### Q9

Redo number Q8 but now with the time series you created on Q7, i.e., the series starting in 2012. Using what `autoplot()` again what happened now? Did the function removed any outliers from the series?

```{r removing outliers cont'd}
# removing outliers with tsclean()
solar2_clean <- tsclean(ts_solar2)
wind2_clean <- tsclean(ts_wind2)

# plotting "cleaned" series
autoplot(solar2_clean)
autoplot(wind2_clean)

```

> Answer: ADD EXPLANATION
