---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 3 - Due date 02/01/24"
author: "Aditi Jackson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
```{r}

```
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Renewable Energy Production and Hydroelectric Power Consumption. Create a data frame structure with these two time series only.

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r installing and loading packages}
#Load/install required package here
#install.packages("forecast")
#install.packages("tseries")
#install.packages("Kendall")
#install.packages("ggplot2")
#install.packages("cowplot")

library("forecast")
library("tseries")
library("Kendall")
library("ggplot2")
library("cowplot")


```


```{r creating time series df}
# checking working directory
getwd()

# loading data using read.csv
renewable_energy_full <- read.csv("./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.csv", header = TRUE, dec = ".", sep=",",stringsAsFactors = TRUE)

# converting Month column to date object using lubridate
renewable_energy_full$Month <- ym(renewable_energy_full$Month)

# renaming Month column to "Date"
colnames(renewable_energy_full)[colnames(renewable_energy_full) == "Month"] <- "Date"

# creating subset of data with Total Renewable Energy Production and Hydroelectric Power Consumption by date
renewable_energy_sub <- renewable_energy_full %>%
  select(
    Date,
    Total.Renewable.Energy.Production,
    Hydroelectric.Power.Consumption)

# Verifying data
head(renewable_energy_sub)

# transforming data into ts object
# start date is Jan 1, 1973
# frequency is 12 (monthly data)
renewable_energy_ts <- ts(renewable_energy_sub,start=c(1973,1),frequency=12)

renewable_energy_ts

```
##Trend Component

### Q1
For each time series, i.e., Renewable Energy Production and Hydroelectric Consumption create three plots: one with time series, one with the ACF and with the PACF. You may use the same code form A2, but I want all the three plots side by side as in a grid.
(Hint: use function `plot_grid()` from the `cowplot` package)

```{r plotting time series, ACF, PACF}

# Total Renewable Energy Production plots

## ts plot
Renewables_tsplot <- ggplot(renewable_energy_sub,aes(x=Date,y=Total.Renewable.Energy.Production))+
  geom_line(color="black")+
  #geom_smooth(method="lm")+
  labs(x="Date",y="Trillion Btu",title="Total Renewable Energy Production")

## acf plot
Renewables_acf <- Acf(renewable_energy_ts[,2],lag.max=40)

## pacf plot
Renewables_pacf <- Pacf(renewable_energy_ts[,2],lag.max=40)

## plotting on same grid
Renewables_allPlots <- plot_grid(Renewables_tsplot,autoplot(Renewables_acf),autoplot(Renewables_pacf),nrow=1)


# Hydroelectric Power Consumption plots

## ts plot
Hydro_tsplot <- ggplot(renewable_energy_sub,aes(x=Date,y=Hydroelectric.Power.Consumption))+
  geom_line(color="black")+
  geom_smooth(method="lm")+
  labs(x="Date",y="Trillion Btu",title="Hydroelectric Power Consumption")

## acf plot
HydroConsump_acf <- Acf(renewable_energy_ts[,3],lag.max=40)

## pacf plot
HydroConsump_pacf <- Pacf(renewable_energy_ts[,3],lag.max=40)

## plotting on same grid
Hydro_allPlots <- plot_grid(Hydro_tsplot,autoplot(HydroConsump_acf),autoplot(HydroConsump_pacf))

```

### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

> COME BACK... If using "i" you might not need gridplot?

### Q3
Use the *lm()* function to fit a linear trend to the two time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.
```{r linear regressions}
# creating vector "t" to represent number of observations
num_obs <- nrow(renewable_energy_sub)
t <- 1:num_obs

# fitting linear regression for Renewable Energy Consumption time series
Renewables_linReg <- lm(Total.Renewable.Energy.Production ~ t, data = renewable_energy_sub)
summary(Renewables_linReg)

# saving coefficients
Renew_beta0 <- Renewables_linReg$coefficients[1]
Renew_beta1 <- Renewables_linReg$coefficients[2]

# ADD INTERPRETATION

# fitting linear regression for Hydropower Production time series
Hydro_linReg <- lm(Hydroelectric.Power.Consumption ~ t, data = renewable_energy_sub)
summary(Hydro_linReg)

# saving coefficients
Hydro_beta0 <- Hydro_linReg$coefficients[1]
Hydro_beta1 <- Hydro_linReg$coefficients[2]

# ADD INTERPRETATION

```
 

### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?
```{r detrend}

# Renewable Energy Production

# de-trending renewable energy production time series 
Renew_detrend <- renewable_energy_sub[,2] - (Renew_beta0+Renew_beta1*t)

# creating data frame in order to plot
df_renew_detrend <- data_frame("date"=renewable_energy_sub$Date,
                               "observed"=renewable_energy_sub[,2],"detrend"=Renew_detrend)

# plotting detrended series
ggplot(df_renew_detrend,aes(x=date))+
  geom_line(aes(y=observed),color="black")+
  geom_line(aes(y=detrend),color="blue")+
  labs(x="Date",y="Observations")


# Hydroelectric Power Consumption
 
# de-trending hydroelectric power consumption time series 
Hydro_detrend <- renewable_energy_sub[,3] - (Hydro_beta0+Hydro_beta1*t)

# creating data frame in order to plot
df_hydro_detrend <- data_frame("date"=renewable_energy_sub$Date,
                               "observed"=renewable_energy_sub[,3],"detrend"=Hydro_detrend)

# plotting detrended series
ggplot(df_hydro_detrend,aes(x=date))+
  geom_line(aes(y=observed),color="black")+
  geom_line(aes(y=detrend),color="blue")+
  labs(x="Date",y="Observations")

# ADD - HOW DO THESE PLOTS DIFFER FROM Q1???

```

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. But mot mandatory. Did the plots change? How?
```{r}


```

## Seasonal Component

Set aside the de-trended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Just by looking at the time series and the acf plots, do the series seem to have a seasonal trend? No need to run any code to answer your question. Just type in you answer below.
> 


### Q7
Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) the two time series. Ask R to print the summary of the regression. Interpret the regression output. From the results which series have a seasonal trend? Do the results match you answer to Q6?
```{r}


```


### Q8
Use the regression coefficients from Q7 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?
```{r}


```


### Q9
Plot ACF and PACF for the deseason series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. nut mot mandatory. Did the plots change? How?
```{r scraps}
# Setting index variables
ncolumns <- ncol(renewable_energy_sub) - 1
nrows <- nrow(renewable_energy_sub)

# Initialize a list to store plots
plot_list <- list()

# Plotting all three time series: Renewable Production & Hydroelectric Consumption
for (i in 1:ncolumns) {
  # Create the plot
  plot <- ggplot(renewable_energy_sub, aes_string(x = "Date", y = colnames(renewable_energy_sub)[(1 + i)])) +
    geom_line(color = "blue") +
    ylab(paste0("TS: ", colnames(renewable_energy_sub)[(1 + i)], sep = ""))
  
# Store the plot in the list
  plot_list[[i]] <- plot
}

# Combine plots using plot_grid
combined_plot <- plot_grid(plotlist = plot_list, ncol = 2)  
print(combined_plot)

# plotting acf and pacf
for (i in 1:ncolumns) {
  par(mfrow = c(1, 2))  # Setting up the layout for ACF/PACF plots
  Acf(renewable_energy_ts[, i], lag.max = 40, main = paste("ACF of TS:", i))
  Pacf(renewable_energy_ts[, i], lag.max = 40, main = paste("PACF of TS:", i))
}

# Store the plot in the list
plot_list[[i]] <- plot

# Combine plots using plot_grid
combined_plot_cf <- plot_grid(plotlist = plot_list, ncol = 1)  # Adjust ncol as needed
print(combined_plot)


```

