---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 2 - Due date 02/25/24"
author: "Aditi Jackson"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here

# installing packages
# install.packages("forecast")
# install.packages("tseries")
# install.packages("dplyr")
# install.packages("lubridate")
# install.packages("ggplot2")

# loading packages
library("forecast")
library("tseries")
library("dplyr")
library("lubridate")
library("ggplot2")

```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2023 Monthly Energy Review. The spreadsheet is ready to be used. You will also find a $.csv$ version of the data "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv". You may use the function $read.table()$ to import the $.csv$ data in R. Or refer to the file "M2_ImportingData_CSV_XLSX.Rmd" in our Lessons folder for functions that are better suited for importing the $.xlsx$.

```{r}
#Importing data set

# checking working directory
getwd()

# loading data using read.csv
# NOTE - R not letting me change my working directory to be my project folder, so
# I moved the file to Assignments > RMD (current wd) as a workaround

renewable_energy_full <- read.csv(
  file="Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.csv",
  header = TRUE, dec = ".", sep=",",stringsAsFactors = TRUE)

# converting Month column to date object using lubridate
renewable_energy_full$Month <- ym(renewable_energy_full$Month)

# renaming Month column to "Date"
colnames(renewable_energy_full)[colnames(renewable_energy_full) == "Month"] <- "Date"

```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r dataframe}

# creating subset of data with Total Biomass Energy Production,
# Total Renewable Energy Production, Hydroelectric Power Consumption by date
renewable_energy_sub <- renewable_energy_full %>%
  select(
    Date,
    Total.Biomass.Energy.Production,
    Total.Renewable.Energy.Production,
    Hydroelectric.Power.Consumption)

# Verifying data
head(renewable_energy_sub)


```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r time series object}

# transforming data into ts object
# start date is Jan 1, 1973
# frequency is 12 (monthly data)
renewable_energy_ts <- ts(renewable_energy_sub,start=c(1973,1),frequency=12)

renewable_energy_ts

```


## Question 3
Compute mean and standard deviation for these three series.

```{r}

# calculating mean and standard deviation for each time series

# mean and sd for biomass
bioMass_mean_ts <- print(mean(renewable_energy_ts[,2]))
bioMass_stDev_ts <- print(sd(renewable_energy_ts[,2]))

# mean and sd for renewable energy
renewEnergy_mean_ts <- print(mean(renewable_energy_ts[,3]))
renewEnergy_stDev_ts <- print(sd(renewable_energy_ts[,3]))

# mean and sd for hydroelectric
Hydro_mean_ts <- print(mean(renewable_energy_ts[,4]))
Hydro_stDev_ts <- print(sd(renewable_energy_ts[,4]))

```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r initial plots}

# plotting Total Biomass Production
ggplot(renewable_energy_sub,aes(x=Date,y=Total.Biomass.Energy.Production))+
  geom_line(color="black")+
  geom_smooth(method="lm")+
  labs(x="Date",y="Trillion Btu",title="Total Biomass Production")

# plotting Total Renewable Energy Production
ggplot(renewable_energy_sub,aes(x=Date,y=Total.Renewable.Energy.Production))+
  geom_line(color="black")+
  geom_smooth(method="lm")+
  labs(x="Date",y="Trillion Btu",title="Total Renewable Energy Production")

# plotting Hydroelectric Power Consumption
ggplot(renewable_energy_sub,aes(x=Date,y=Hydroelectric.Power.Consumption))+
  geom_line(color="black")+
  geom_smooth(method="lm")+
  labs(x="Date",y="Trillion Btu",title="Hydroelectric Power Consumption")

```
Interpretation of Time Series Plots
> Total Biomass Production: The plot shows an upward trend in production over time. We can see this from the highly positive slope of the trend line in blue. There also seems to be some seasonanality in the plot as similar cyclical patterns seem to repeat.

> Total Renewable Energy Production: Similarly to biomass production, there appears to be a positive trend in renewable energy production over time. The plot is very similar to that of the Biomass plot, which is unsurprising given the increasing demand alternatives to fossil energy. Like above, there seems to be some seasonality.

> Hydroelectric Power Consumption: There is a slight downward trend in hydroelectric power consumption over time. This is interesting and stands in constrast to Biomass and Renewables production. This downward trend could be because consumers are switching to alternative forms of renewable energy as production / availability increases.

## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r correlation}

# NOTE: Used chatGPT to determine which function to use
# to calculate correlation between all three time series

cor_matrix <- 
  cor(renewable_energy_sub[,c("Total.Biomass.Energy.Production",
                              "Total.Renewable.Energy.Production",
                              "Hydroelectric.Power.Consumption")])

print(cor_matrix)

```
Answer: 
> It appears that biomass production and renewable energy production are highly correlated with a correlation coefficient of 0.97. This does not seem that surprising since both variables relate to the production of alternative forms of energy (non-fossil). Meanwhile, hydroelectric power consumption does not appear to be correlated with either biomass or renewable energy production based on the fact the correlation coefficients are > 0.1.

## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r acf and pacf}

BiomassProd_acf <- Acf(renewable_energy_ts[,2],lag.max=40)

RenewablesProd_acf <- Acf(renewable_energy_ts[,3],lag.max=40)

HydroConsump_acf <- Acf(renewable_energy_ts[,4],lag.max=40)

```
Answer:
> The biomass and renewable energy ACF plots behave similarly,but the hydro power plot of ACF looks very different. The biomass and renewable energy plots show significant correlation within each dataset. ACF appears positive for all lags for these two time series. For hydro power, some autocorrelation is present, but the amplitudes don't appear as big compared to the first two time series, suggesting potentially less correlation between variables.

## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}

BiomassProd_pacf=Pacf(renewable_energy_ts[,2],lag.max=40)

RenewablesProd_pacf=Pacf(renewable_energy_ts[,3],lag.max=40)

HydroConsump_pacf=Pacf(renewable_energy_ts[,4],lag.max=40)

```
Answer:
> The PACF plots show us which lags are most significant within each time series. In other words, which instances within a specific variable are most correlated with each other. From these plots we should consider adding lags that breach the blue dotted line to our model as they may help us better explain trends / construct more accurate forecasts. 
