---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 2 - Due date 02/25/24"
author: "Aditi Jackson"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here

# installing packages
# install.packages("forecast")
# install.packages("tseries")
# install.packages("dplyr")
# install.packages("lubridate")
# install.packages("ggplot2")

# loading packages
library("forecast")
library("tseries")
library("dplyr")
library("lubridate")
library("ggplot2")

```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2023 Monthly Energy Review. The spreadsheet is ready to be used. You will also find a $.csv$ version of the data "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv". You may use the function $read.table()$ to import the $.csv$ data in R. Or refer to the file "M2_ImportingData_CSV_XLSX.Rmd" in our Lessons folder for functions that are better suited for importing the $.xlsx$.

```{r}
#Importing data set

# checking working directory
getwd()

# loading data using read.csv; NOTE - R was not letting me change my working directory to be my project folder, so I moved the file to Assignments > RMD (current wd) as a workaround
renewable_energy_full <- read.csv(file="Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.csv", header = TRUE, dec = ".", sep=",",stringsAsFactors = TRUE)

# converting Month column to date object using lubridate
renewable_energy_full$Month <- ym(renewable_energy_full$Month)

# renaming Month column to "Date"
colnames(renewable_energy_full)[colnames(renewable_energy_full) == "Month"] <- "Date"

```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r dataframe}

# creating subset of data with Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption by date
renewable_energy_sub <- renewable_energy_full %>%
  select(Date,Total.Biomass.Energy.Production,Total.Renewable.Energy.Production,Hydroelectric.Power.Consumption)

# Verifying data
head(renewable_energy_sub)


```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r time series object}

# transforming data into ts object
renewable_energy_ts <- ts(renewable_energy_sub,start=c(1973,1),frequency=12)

```


## Question 3
Compute mean and standard deviation for these three series.

```{r}

# calculating mean and standard deviation for time series
mean_ts <- mean(renewable_energy_ts)
stDev_ts <- sd(renewable_energy_ts)


# mean / standard deviation for each separately
renewable_energy_stats <- renewable_energy_sub %>%
  summarize(Mean.Biomass.Prod = mean(Total.Biomass.Energy.Production),
            Mean.Renew.Energy.Prod = mean(Total.Renewable.Energy.Production),
            Mean.Hydro.Cons = mean(Hydroelectric.Power.Consumption),
            SD.Biomass.Prod = sd(Total.Biomass.Energy.Production),
            SD.Renew.Energy.Prod = sd(Total.Renewable.Energy.Production),
            SD.Hydro.Cons = sd(Hydroelectric.Power.Consumption))

```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r initial plots}
autoplot(renewable_energy_ts) +
  xlab("Time") +
  ylab("Trillion Btu")

# FIX GRAPH, ADD INTERPRETATION

```


## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r correlation}



```


## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}

BiomassProd_acf=Acf(renewable_energy_ts[,1],lag.max=40, type="correlation", plot=TRUE)

RenewablesProd_acf=Acf(renewable_energy_ts[,2],lag.max=40, type="correlation", plot=TRUE)

HydroConsump_acf=Acf(renewable_energy_ts[,3],lag.max=40, type="correlation", plot=TRUE)
```


## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}

BiomassProd_pacf=Pacf(renewable_energy_ts[,1],lag.max=40, type="correlation", plot=TRUE)

RenewablesProd_pacf=Pacf(renewable_energy_ts[,2],lag.max=40, type="correlation", plot=TRUE)

HydroConsump_pacf=Pacf(renewable_energy_ts[,3],lag.max=40, type="correlation", plot=TRUE)

```


